Unifying intra- and inter-specific variation in tropical tree mortality

### 2017-06-16 ms/ms.Rnw
Changes to sub sampling

To validate this model, we fit a series of models with increasing complexity (eq. \ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_unique)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken over a 15 yr period (Fig. \ref{fig1}A). We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across five iterative stages of model development. At each stage, the best model from the previous stage was taken as input for the next stage.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive skill (Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, (\frac{\rho}{\rho_c})^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$, modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. Full distributions for the final model are shown in Figs. S3\textendash S5.

>>>>

To validate this model, we fit a series of models with increasing complexity (eq. \ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken over a 15 yr period (Fig. \ref{fig1}A). We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across five iterative stages of model development. At each stage, the best model from the previous stage was taken as input for the next stage.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive skill (Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, (\frac{\rho}{\rho_c})^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$, modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. Full distributions for the final model are shown in Figs. S3\textendash S5.
